From Reversible Programs to \\ Univalent Universes and Back

### 2021-11-11 Pi/paper/reversible1.tex
Final revisions

All of this is only half the story, however. A sound semantics for $\PiLang$ using weak rig groupoids was established
by~\citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this semantics, \emph{coherence
conditions} for symmetric rig groupoids that identify different syntactic representations of the same
permutation~\cite{laplazaCoherenceDistributivity1972,caretteComputingSemiringsWeak2016}, were collected in a second
level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for
appropriate $c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same
bijection. For example, we have the following level-2 combinators dealing with associativity:
\medskip
\leveltwoblockone{}

\noindent The full set of level-2 combinators is large; the remaining combinators are listed in~\cref{app:leveltwo} in
the appendix of the extended version~\cite{choudhurySymmetriesReversibleProgramming2021}, and also in the accompanying
Agda formalisation~\cite{choudhuryArtifactSymmetriesReversible2021}.

>>>>

All of this is only half the story, however. A sound semantics for $\PiLang$ using weak rig groupoids was established
by~\citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this semantics, \emph{coherence
conditions} for symmetric rig groupoids that identify different syntactic representations of the same
permutation~\cite{laplazaCoherenceDistributivity1972}, were collected in a second level of $\PiLang$ syntax as level-2
combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate $c_1$ and $c_2$ of the same
level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For example, we have the
following level-2 combinators dealing with associativity:
\medskip
\leveltwoblockone{}

\noindent The full set of level-2 combinators is large; the remaining combinators are listed in~\cref{app:leveltwo} in
the extended version~\cite{choudhurySymmetriesReversibleProgramming2021}, and also in the accompanying Agda
formalisation~\cite{choudhuryArtifactSymmetriesReversible2021}.

### 2021-11-11 Pi/paper/ufin.tex
Final revisions

In categorical language, the target for the semantics in the previous section is the category of finite sets and
functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
$\BFin = \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and
finite products $(\unit, \times)$ and in $\BFin$ these restrict to additive and multiplicative symmetric monoidal
structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
commutative rig of natural numbers $\Nat$~\cite{baezHIGHERDIMENSIONALALGEBRA2010}.

The semantics interprets types of $\PiLang$ as objects in $\BFin$, 1-combinators as isomorphisms, and for every pair of
1-combinators related by a 2-combinator, their interpretations in~$\BFin$ are equal. $\BFin$ is a 1-groupoid, where the
collection of isomorphisms is a set. We can evaluate two bijections on their inputs to decide if they are equal, and
there is no higher cell witnessing the equality of two isomorphisms. To be able to establish completeness for $\PiLang$,
we want a witness for this equality, so that we can quote back to the syntax and produce a 2-combinator witnessing the
equality of the corresponding 1-combinator.

@@

% Fourth - explain the tool we use for fixing it
Every type in HoTT is a weak $\infty$-groupoid whose points are the terms of the type, and the (iterated) identity type
gives the (higher) morphisms. The groupoid we are interested in has types as points, type equivalences for 1-cells, and
higher homotopies for higher cells. (See~\cref{app:grpdexample} for an example on a 3-element set.) This is the
groupoid structure for the universe type $\UU$, since the identity type on types can be characterised as type
equivalences (by \emph{univalence}). But, we only want to carve out a \emph{subuniverse} of \emph{finite types}, still
satisfying univalence, to get the groupoid structure. In this section, we formally define \emph{univalent subuniverses},
and proceed to construct the particular instance for finite types, $\UFin$ (\cref{def:ufin}).

@@

Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action on
the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to the
universe, such as $A \to \UU$, which is an $A$-indexed family of groupoids. For a type family $P : A \to \UU$ and a
point $x : A$, the type $P(x)$ is the fiber over $x$.  The $\term{transport}$ operation, named
$\term{transport}$, lifts paths in the indexing type to functions between fibers.
\begin{gather*}
  \term{ap}_{f} : \dfun*{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
  \qquad
  \term{transport}_{P} : \dfun*{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
\end{gather*}

@@

While $\UFin$ has \emph{all} the finite types, we are also interested in constructing a subuniverse of finite types of a
specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$, for any type $T : \UU$.
Characterisations of univalent fibrations using the $\BAut$ construction have also been studied
by~\citet{christensenCharacterizationUnivalentFibrations2015}.

\begin{definition}[$\BAut$]
  The predicate $P(X) \defeq \Trunc[-1]{X \id T}$ picks out exactly those types that are merely equal to $T$, and this
  generates the subuniverse
  \[
    \BAut[T] \defeq \Sub{T}.
  \]
  We write $T_0 \defeq (T, \trunc{\refl_{T}})$ for the image of the inclusion of $T$ in $\BAut[T]$.
\end{definition}

@@

\begin{definition}[${\UFin[n]}$]
  For any $n : \Nat$, we define $\UFin[n] \defeq \BAut[\Fin[n]]$ to be the univalent subuniverse of $n$-element sets.
  Note that, $\UFin$ can be equivalently seen as the collection of all types of finite cardinality, that is,
  $\UFin \eqv \dsum*{n:\Nat}{\UFin[n]}$.
\end{definition}

Since $\BAut[T]$ is a univalent subuniverse, we can characterise its path space. The intuition is that $\BAut[T]$ only
has one point $T_0$, and 1-paths $T_0 \id T_0$, that is, loops, and higher paths between these loops. The
type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$, is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
the group of automorphisms of $T$.

@@

\begin{theorem}
  $\UFin[n]$ is a pointed, connected, 1-groupoid for every $n:\Nat$, and \( \loopspace[\UFin[n],F_{n}] \eqv
  \Aut[\Fin[n]] \). $\UFin$ is a 1-groupoid with connected components for every $n:\Nat$.
\end{theorem}

We have shown that loops in $\UFin$ exactly encode the automorphism group $\Aut[\Fin[n]]$ for every~$n$. This is a
general technique called \emph{delooping}, where a group can be identified with a 1-object groupoid, that is, a
0-connected type in HoTT. This technique also allows defining higher groups~\cite{buchholtzHigherGroupsHomotopy2018}.
The loopspace of a pointed type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element,
path composition for the group multiplication, and path inverse for the group inverse operation. The group axioms are
given by the higher paths corresponding to groupoid laws.

@@

\begin{proposition}
  For any $n, m : \Nat$, and for any types $X, Y, Z$,
  \begin{gather*}
    \begin{aligned}
      \Fin[0]                & \eqv \bot                  \\
      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]           \\[1.5ex]
      \bot \sqcup X          & \eqv X                     \\
      X \sqcup \bot          & \eqv X                     \\
      (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
      X \sqcup Y             & \eqv Y \sqcup X            \\
      X \times \bot          & \eqv \bot
    \end{aligned}
    \qquad
    \begin{aligned}
      \Fin[1]                & \eqv \top                             \\
      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]                      \\[1.5ex]
      \top \times X          & \eqv X                                \\
      X \times \top          & \eqv X                                \\
      (X \times Y) \times Z  & \eqv X \times (Y \times Z)            \\
      X \times Y             & \eqv Y \times X                       \\
      X \times (Y \sqcup Z)  & \eqv (X \times Y) \sqcup (X \times Z)
    \end{aligned}
  \end{gather*}
\end{proposition}

\begin{theorem}
  $\UFin$ has two symmetric monoidal structures, the additive and multiplicative ones, given by $(F_0, \sqcup)$ and
  $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and the braiding
  isomorphism $\mathcal{B}_{X,Y}$ upto 1-paths in $\UFin$. These isomorphisms satisfy the Mac Lane coherence conditions
  for symmetric monoidal categories~\cite{maclaneNaturalAssociativityCommutativity1963}, that is, the triangle,
  pentagon, and hexagon identities, and the symmetry of the braiding, upto 2-paths in $\UFin$. The multiplicative
  structure distributes over the additive structure and satisfies the Laplaza coherence conditions for rig
  categories~\cite{laplazaCoherenceDistributivity1972}.
\end{theorem}

>>>>

In categorical language, the target for the semantics in the previous section is the category of finite sets and
functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
$\BFin = \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and
finite products $(\unit, \times)$, and in $\BFin$, these restrict to additive and multiplicative symmetric monoidal
structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
commutative rig of natural numbers $\Nat$~\cite{baezHIGHERDIMENSIONALALGEBRA2010}.

The semantics interprets types of $\PiLang$ as objects in $\BFin$, 1-combinators as isomorphisms, and for every pair of
1-combinators related by a 2-combinator, their interpretations in~$\BFin$ are equal. $\BFin$ is a (strict) 1-groupoid
where isomorphisms are identified upto extensional equality. We can evaluate two bijections on their inputs to decide if
they are equal, and there is no higher cell witnessing the equality of two isomorphisms. To be able to establish
completeness for $\PiLang$, we want a witness for this equality, so that we can quote back to the syntax and produce a
2-combinator witnessing the equality of the corresponding 1-combinator.

@@

% Fourth - explain the tool we use for fixing it
Every type in HoTT is a weak $\infty$-groupoid whose points are the terms of the type, and the (iterated) identity type
gives the (higher) morphisms. The groupoid we are interested in has types as points, type equivalences for 1-cells, and
higher homotopies for higher cells. (See~\cref{app:grpdexample} for an example on a 3-element set.) This is the groupoid
structure for the universe type $\UU$, since the identity type on types can be characterised as type equivalences (by
\emph{univalence}). But, we only want to carve out a \emph{subuniverse} of \emph{finite types}, still satisfying
univalence, to get the groupoid structure. In this section, we formally define \emph{univalent subuniverses}, and
proceed to construct the particular instance for finite types, $\UFin$ (see~\cref{def:ufin}).

@@

Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action on
the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to the
universe, such as $A \to \UU$, which is an $A$-indexed family of groupoids. For a type family $P : A \to \UU$ and a
point $x : A$, the type $P(x)$ is the fiber over $x$.  The $\term{transport}$ operation lifts paths in the indexing type
to functions between fibers.
\begin{gather*}
  \term{ap}_{f} : \dfun*{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
  \qquad
  \term{transport}_{P} : \dfun*{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
\end{gather*}

@@

While $\UFin$ has \emph{all} the finite types, we are also interested in constructing a subuniverse of finite types of a
\emph{specified} cardinality. To do so, we will start with the subuniverse $\BAut[T]$, for any type $T : \UU$.
Characterisations of univalent fibrations using the $\BAut$ construction have also been studied
by~\citet{christensenCharacterizationUnivalentFibrations2015}.

\begin{definition}[$\BAut$]
  The predicate $P(X) \defeq \Trunc[-1]{X \eqv T}$, or equivalently, $\Trunc[-1]{X \id T}$, picks out exactly those
  types that are merely equivalent to $T$, and this generates the subuniverse
  \[
    \BAut[T] \defeq \Sub{T}.
  \]
  We write $T_0 \defeq (T, \trunc{\refl_{T}})$ for the image of the inclusion of $T$ in $\BAut[T]$.
\end{definition}

@@

\begin{definition}[${\UFin[n]}$]
  For any $n : \Nat$, we define $\UFin[n] \defeq \BAut[\Fin[n]]$ to be the univalent subuniverse of $n$-element sets.
  Note that, $\UFin$ can be equivalently seen as the collection of all finite types, that is, $\UFin \eqv
    \dsum*{n:\Nat}{\UFin[n]}$.
\end{definition}

Since $\BAut[T]$ is a univalent subuniverse, we can characterise its path space. The intuition is that $\BAut[T]$ only
has one point or 0-cell $T_0$, and 1-paths $T_0 \id T_0$, that is, loops, and higher paths between these loops. The type
of loops on $T_0$, that is, $\loopspace[\BAut[T],T_{0}]$, is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which
is the group of automorphisms of $T$.

@@

\begin{theorem}
  \leavevmode
  \begin{enumerate}
    \item $\UFin[n]$ is a pointed, connected, 1-groupoid for every $n:\Nat$, and \( {\loopspace[\UFin[n],F_{n}] \eqv
          \Aut[\Fin[n]]} \).
    \item $\UFin$ is a 1-groupoid with connected components for every $n:\Nat$.
  \end{enumerate}
\end{theorem}

We have shown that loops in $\UFin$ exactly encode the automorphism group $\Aut[\Fin[n]]$ for every~$n$. This is a
general technique called \emph{delooping}, where a group can be identified with a 1-object groupoid, that is, a
0-connected type in HoTT. This technique also allows defining higher groups~\cite{buchholtzHigherGroupsHomotopy2018}.
The loopspace of a pointed type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element,
path composition for the group multiplication, and path inverse for the group inverse. The group axioms follow from the
higher paths corresponding to groupoid laws.

@@

\begin{proposition}
  For any $n, m : \Nat$, and for any types $X, Y, Z$,
  \begin{gather*}
    \begin{aligned}
      \Fin[0]                & \eqv \bot                  \\
      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]           \\[1.5ex]
      \bot \sqcup X          & \eqv X                     \\
      X \sqcup \bot          & \eqv X                     \\
      (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
      X \sqcup Y             & \eqv Y \sqcup X            \\
      X \times \bot          & \eqv \bot
    \end{aligned}
    \qquad
    \begin{aligned}
      \Fin[1]                & \eqv \top                             \\
      \Fin[n] \times \Fin[m] & \eqv \Fin[n \cdot m]                  \\[1.5ex]
      \top \times X          & \eqv X                                \\
      X \times \top          & \eqv X                                \\
      (X \times Y) \times Z  & \eqv X \times (Y \times Z)            \\
      X \times Y             & \eqv Y \times X                       \\
      X \times (Y \sqcup Z)  & \eqv (X \times Y) \sqcup (X \times Z)
    \end{aligned}
  \end{gather*}
\end{proposition}

\begin{theorem}
  $\UFin$ has two symmetric monoidal structures, the additive and multiplicative ones, given by $(F_0, \sqcup)$ and
  $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and the braiding
  isomorphism $\mathcal{B}_{X,Y}$, upto 1-paths in $\UFin$. These isomorphisms satisfy the Mac Lane coherence conditions
  for symmetric monoidal categories~\cite{maclaneNaturalAssociativityCommutativity1963}, that is, the triangle,
  pentagon, and hexagon identities, and the symmetry of the braiding, upto 2-paths in $\UFin$. The multiplicative
  structure distributes over the additive structure and satisfies the Laplaza coherence conditions for rig
  categories~\cite{laplazaCoherenceDistributivity1972}.
\end{theorem}
